Parallel Level-Set Methods on Adaptive Tree-Based Grids

### 2016-02-17 paper/tex/introduction.tex
Some edits on paper

One of the main challenges in parallelizing algorithms on adaptive grids is
handling the grid itself.
One option is to replicate the entire grid on each process and to employ serial
ordering techniques, as originally implemented by the \texttt{deal.II} library
\cite{Bangerth;Hartmann;Kanschat:07:deal.II----a-General}, or serial
graph partitioners such as METIS \cite{KarypisKumar95}.
%
% off-the-shelf graph partitioners, e.g.\ ParMetis
% \cite{Karypis;Kumar:98:A-parallel-algorithm} or Zoltan
% \cite{Boman;Catalyurek;Chevalier;etal:12:The-Zoltan-and-Isorr}, for load
% balancing and domain decomposition.
% For instance, this was the approach originally taken by the \texttt{deal.II}
% library \cite{Bangerth;Hartmann;Kanschat:07:deal.II----a-General}.
%
This approach, however, is only scalable to a few hundred processes at best and
is limited by the size of the grid itself that can fit in memory.
Even though parallel general-purpose partitioners have since been popularized
\cite{Karypis;Kumar:98:A-parallel-algorithm,
Boman;Catalyurek;Chevalier;etal:12:The-Zoltan-and-Isorr} and
the scalability of partitioning algorithms for unstructured grids
has been improved (see e.g.\ \cite{SahniZhouShephardEtAl09}),
% the use of a general-purpose graph partitioner
their use
adds extra overhead that can limit the overall scalability.
Interestingly, tree-based grids have a nice spatial ordering that naturally
leads to the concept of space-filling curves (SFCs) and can be efficiently
exploited for parallel load balancing
\cite{Aluru;Sevilgen:97:Parallel-domain-deco, GriebelZumbusch99,
      Campbell;Devine;Flaherty;etal:03:Dynamic-octree-load-}.

@@

In implementing the numerical solver, we make use of the popular \texttt{PETSc}
\cite{Balay;Abhyankar;Adams;etal:14:PETSc-Web-page} library for linear algebra
and its parallel primitives, such as parallel ghosted vector and scatter/gather
operations, which simplifies the implementation.

@@

In many applications, it is desirable that the level-set function has the signed-distance property, i.e., $|\nabla \phi| = 1$. Generally, there are two approaches to enforce this property, either by solving the pseudo-time transient reinitialization equation \cite{Sussman;Smereka;Osher:94:A-Level-Set-Approach, Osher;Fedkiw:01:Level-Set-Methods:-A}
\ben
\phi_\tau + S(\phi_0)\left(|\nabla \phi| - 1\right) = 0,
\een
or by solving the Eikonal equation
\ben
F(x)|\nabla\phi| = 1
\een 
with constant speed function $F(x) \equiv 1$. The transient reinitialization equation can be solved using explicit finite differences and thus can easily be parallelized in a domain decomposition approach. Moreover, only a few iterations may be needed if the signed-distance property is only required close to the interface \cite{Min;Gibou:07:A-second-order-accur}. This is the approach we have chosen in this paper. However, if the signed-distance property is required in the entire domain, solving the Eikonal equation is more computationally efficient. Unfortunately, the most popular algorithm for solving the Eikonal equation, i.e.\ the Fast Marching Method \cite{Sethian:96:A-Fast-Marching-Leve,Sethian:99:Level-set-methods-an}, is inherently sequential due to causal relationship between grid points and cannot be easily parallelized.
The Fast Sweeping Method (FSM) \cite{Zhao:05:A-fast-sweeping-meth} is an
alternative for solving the Eikonal equation iteratively.
The FSM can be more computationally efficient for simple choices of speed
function, e.g.\ as in this context, and for simple interfaces.
Moreover, FSM has more potential for parallelization compared to the FMM.

>>>>

One of the main challenges in parallelizing algorithms on adaptive grids is
handling the grid itself.
One option is to replicate the entire grid on each process and to employ serial
ordering techniques, as implemented in an earlier version of the \texttt{deal.II} library
\cite{Bangerth;Hartmann;Kanschat:07:deal.II----a-General}, or to use serial
graph partitioners such as METIS \cite{KarypisKumar95}.
%
% off-the-shelf graph partitioners, e.g.\ ParMetis
% \cite{Karypis;Kumar:98:A-parallel-algorithm} or Zoltan
% \cite{Boman;Catalyurek;Chevalier;etal:12:The-Zoltan-and-Isorr}, for load
% balancing and domain decomposition.
% For instance, this was the approach originally taken by the \texttt{deal.II}
% library \cite{Bangerth;Hartmann;Kanschat:07:deal.II----a-General}.
%
This approach, however, is only scalable to a few hundred processes at best and
is limited by the size of the grid itself that can fit in memory.
Even though parallel general-purpose partitioners have since been popularized
\cite{Karypis;Kumar:98:A-parallel-algorithm,
Boman;Catalyurek;Chevalier;etal:12:The-Zoltan-and-Isorr} and
the scalability of partitioning algorithms for unstructured grids
has been improved (see e.g.\ \cite{SahniZhouShephardEtAl09}),
% the use of a general-purpose graph partitioner
their use
adds extra overhead that can limit the overall scalability.
Refining a grid consisting of multiple trees using recursive coordinate
bisection has been implemented in the SIERRA framework \cite{StewartEdwards04}.
Interestingly, tree-based grids have a nice spatial ordering that naturally
leads to the concept of space-filling curves (SFCs) and can be efficiently
exploited for parallel load balancing
\cite{Aluru;Sevilgen:97:Parallel-domain-deco, GriebelZumbusch99,
      Campbell;Devine;Flaherty;etal:03:Dynamic-octree-load-}.

@@

@@

In many applications, it is desirable that the level-set function has the signed-distance property, i.e., $|\nabla \phi| = 1$. Generally, there are two approaches to enforce this property, either by solving the pseudo-time transient reinitialization equation \cite{Sussman;Smereka;Osher:94:A-Level-Set-Approach, Osher;Fedkiw:01:Level-Set-Methods:-A}
\ben
\phi_\tau + S(\phi_0)\left(|\nabla \phi| - 1\right) = 0,
\een
or by solving the Eikonal equation
\ben
F(x)|\nabla\phi| = 1
\een 
with constant speed function $F(x) \equiv 1$. The transient reinitialization equation can be solved using explicit finite differences and thus can easily be parallelized in a domain decomposition approach. Moreover, only a few iterations may be needed if the signed-distance property is only required close to the interface \cite{Min;Gibou:07:A-second-order-accur}. This is the approach we have chosen in this paper. However, if the signed-distance property is required in the entire domain, solving the Eikonal equation is more computationally efficient. Unfortunately, the most popular algorithm for solving the Eikonal equation, i.e.\ the Fast Marching Method (FMM) \cite{Sethian:96:A-Fast-Marching-Leve,Sethian:99:Level-set-methods-an}, is inherently sequential due to causal relationship between grid points and cannot be easily parallelized.
The Fast Sweeping Method (FSM) \cite{Zhao:05:A-fast-sweeping-meth} is an
alternative for solving the Eikonal equation iteratively.
The FSM can be more computationally efficient for simple choices of speed
function, e.g.\ as in this context, and for simple interfaces.
Moreover, FSM has more potential for parallelization compared to the FMM.

### 2016-02-17 paper/tex/parallel_algorithms.tex
fixed a minor error in figure caption
Some edits on paper

In \texttt{p4est} the adaptive grid is represented as a non-overlapping collection of trees that are rooted in individual cells of a common coarse grid (cf.\ Figure \ref{fig:p4est_zcurve}). This common coarse grid, which we will refer to as the ``macromesh'', can in general be an unstructured quadrilateral mesh, in two spatial dimensions, or hexahedral mesh, in three spatial dimensions.
In this article, however, it is sufficient to limit discussions to simple
uniform Cartesian macromeshes.
Moreover, it is implicitly assumed that the macromesh is small enough that it
can be entirely replicated on all processes.
For instance, in many of the applications that we are considering in this paper the macromesh is simply a single cell.
\texttt{p4est} allows for arbitrary refinement and coarsening criteria through defining callback functions.
In this article the refinement criteria is chosen based on the distance of individual cells to the interface.
Specifically, a cell $C$ is marked for refinement if
\be
\min_{v \in V(C)} |\phi (v)| \le \frac{L D}{2},
\label{eq:refine}
\ee
where $V(C)$ denotes the set of all vertices of cell $C$, $L$ denotes the
Lipschitz constant of the level-set function, and $D$ denotes the diagonal size
of cell $C$.
Conversely, an existing cell is marked for coarsening if
\be
\min_{v \in V(C)} |\phi (v)| > L D.
\label{eq:coarsen}
\ee
We refer to section 3.2 of
\cite{Burstedde;Wilcox;Ghattas:11:p4est:-Scalable-Algo} for details on the
parallel refinement and coarsening algorithms implemented in \texttt{p4est}.

Once the grid is adapted to the interface, it must be partitioned to ensure
load balancing across processes.
This is achieved by constructing a Z-curve that traverses the leaves of all
trees in order of the tree index (cf.\ Figure \ref{fig:p4est_zcurve}).
A Z-curve is a Space Filling Curve (SFC) with the important property that cells
with close Z-indices are also geometrically close (on average) in the physical
domain.
This is beneficial since it leads to both a reduction in MPI communications and
improvements of the cache performance of several algorithms such as
interpolation and finite difference calculations.
For more details on parallel partitioning in \texttt{p4est} one may refer to
section 3.3 of \cite{Burstedde;Wilcox;Ghattas:11:p4est:-Scalable-Algo}.
\begin{figure}[hbtp]
\begin{center}
\includegraphics[width = \columnwidth]{figures/p4est_zcurve.pdf}
\caption{Left: a ``forest'' made up of two trees $T_0$ and $T_1$. Parallel
partitioning is achieved by first constructing a Z-curve starting at cell ``S''
and ending at cell ``E''.  Next, the one dimensional curve is split up among
processes either uniformly or by assigning different weights to cells.  Here
processes are represented via different colors.  Note how using the Z-curve
naturally leads to clustering of most cells in each domain.  Right: schematic of
a tree data structure representing this forest and its partitioning.}
\label{fig:p4est_zcurve}
\end{center}
\end{figure}
Aside from grid manipulation and partitioning, we use two additional features
of \texttt{p4est}, namely the generation of ghost layer cells and the creation
of a globally unique node indexing.
These algorithms are detailed in sections 3.5 and 3.6 of
\cite{Burstedde;Wilcox;Ghattas:11:p4est:-Scalable-Algo}.
We have specifically extended the latter algorithm such that it can be applied
to a non-graded refinement pattern.
This is important because we can entirely skip the 2:1 balance function, which
was shown to be one of the most time consuming parts of grid adaptation
in \texttt{p4est}
\cite{Burstedde;Wilcox;Ghattas:11:p4est:-Scalable-Algo}.
% (it has since been optimized \cite{IsaacBursteddeGhattas12}).

Finally, in \texttt{p4est} trees are linearized, i.e.\ only the leaves are explicitly stored. However, explicit knowledge of the hierarchal structure of the tree is greatly beneficial in several algorithms, e.g. in search operations needed for the interpolation algorithm. Thus, we introduce a simple reconstruction algorithm that recreates a local representation of the entire ``forest'' that is only adapted to local cells and, potentially, the ghost layer. This approach is similar to the ideas introduced in \cite{Bangerth;Burstedde;Heister;etal:11:Algorithms-and-data-} and our tests show that in a typical application they amount to less that $1\%$ of the entire runtime. Algorithm \ref{alg:reconstruction} illustrates how this reconstruction is performed. Given a forest and a layer of ghost cells from \texttt{p4est}, the algorithm generates a local representation of the forest by recursively refining from the root until reaching the same level and location of all leaves in the local forest and the ghost layer. Note that algorithm \ref{alg:reconstruction} does not involve any communication and is load balanced provided that the initial forest is balanced. Figure \ref{fig:reconstruction} illustrates an example where Algorithm \ref{alg:reconstruction} is applied. Note how each process has independently generated a local representation of the forest that is refined to match the same leaves as in the global forest and ghost layer.
\begin{figure}[htbp]
\begin{center}
\includegraphics[width = \textwidth]{figures/reconstruct.pdf}
\end{center}
\caption{Left: a forest refined close to an interface and partitioned among four processes, as indicated by colors. Right: each process independently recreates a local forest that is refined to match the local grid and is as coarse as possible elsewhere. Note that empty cells are fictitious, i.e.\ they are only required to generate the hierarchal structure and are not matched by any corresponding cell in the global forest.}
\label{fig:reconstruction}
\end{figure}
\begin{algorithm}[htbp]
\caption{$H \gets \texttt{Reconstruct (}G\texttt{)}$: 
Construction of the local tree hierarchy, $H$, from the parallel grid, $G$, supplied by \texttt{p4est}. The algorithm starts at the coarsest level, i.e. the macromesh, and recursively splits the each cell to match the finest local and ghost cells generated by \texttt{p4est}.}
\begin{algorithmic}[1]
\State $H \gets G.\texttt{macromesh()}$ \Comment{start with the coarsest  grid}
\For {$\mathit{tr} \ein G.\texttt{local\_trees()}$} \Comment{update the hierarchy to match local cells}
	\For {$c \ein \mathit{tr}.\texttt{local\_cells()}$}
		\State $H.\texttt{update\_tree(} \mathit{tr}, c \texttt{)}$
	\EndFor
\EndFor

@@

\subsection{Interpolation and semi-Lagrangian methods} \label{sec:interpolation_semilagrangian}
As indicated earlier, we use the semi-Lagrangian method to solve equation \eqref{eq:ls} when the velocity field is externally generated, i.e.\ when it does not depend explicitly on the level-set function itself. Let us rewrite equation \eqref{eq:ls} along the characteristic curve $\underline{\mathbf{X}}(t)$ as:
\be
\left\{
\begin{array}{rcl}
\dfrac{\text{d} \underline{\mathbf{X}}}{\text{d} t} &=& \underline{\mbf{u}}, \\ [3ex]
\dfrac{\text{d} \phi(\underline{\mathbf{X}}(t), t)}{\text{d} t} &=& 0.
\end{array}
\right.
\label{eq:sl}
\ee
The semi-Lagrangian method integrates equations \eqref{eq:sl} backward in time,
i.e.\ starting from the grid $G^{n+1}$ (computed in iteratively as explained
later on), we simply write $\phi^{n+1}(\underline{\mathbf{X}}^{n+1}) =
\phi(\underline{\mathbf{X}}(t^{n+1}), t^{n+1}) =
\phi(\underline{\mathbf{X}}(t^n), t^n) = \phi^n(\underline{\mathbf{X}}_d)$.
Here, the characteristic curves are chosen such that $\underline{\mathbf{X}}(t^{n+1})$ are simply the coordinates of grids points of $G^{n+1}$, and $\underline{\mathbf{X}}_d$ are the departure points, which are computed using the second-order midpoint method \cite{Min;Gibou:07:A-second-order-accur}:
\bea
\underline{\mathbf{X}}^\star &=& \underline{\mathbf{X}}^{n+1} - \frac{\Delta t}{2} \underline{\mbf{u}}^{n}(\underline{\mathbf{X}}^n),	   \label{eq:xstar}      \\
\underline{\mathbf{X}}_d     &=& \underline{\mathbf{X}}^{n+1} - \Delta t \underline{\mbf{u}}^{n+\frac{1}{2}}(\underline{\mathbf{X}}^\star), \label{eq:xdeparture}
\eea
where $\underline{\mbf{u}}^{n+\frac{1}{2}}$ is obtained via extrapolation from previous times, i.e.:
\be
\underline{\mbf{u}}^{n+\frac{1}{2}} = \frac{3}{2} \underline{\mbf{u}}^n - \frac{1}{2}\underline{\mbf{u}}^{n-1}. \label{eq:vn_p_half}
\ee
Note that all values at the intermediate point, $\underline{\mathbf{X}}^\star$,
and departure point, $\underline{\mathbf{X}}_d$, must be calculated via
interpolation from the previous grids $G^{n}$ and $G^{n-1}$.
Here, we use the stabilized second-order interpolation for
$\phi(\underline{\mathbf{X}}_d)$ and the multi-linear interpolation for
$\underline{\mbf{u}}^{n+\frac{1}{2}}(\underline{\mathbf{X}}^\star)$
\cite{Min;Gibou:07:A-second-order-accur}.
Although parallelization of the interpolation process on a shared-memory
machine is trivial, the same cannot be said for distributed-memory machines.
In fact, the parallel interpolation given in Algorithm \ref{alg:interpolation}
is probably the most important contribution of this article since this
procedure, which is trivial on uniform grids, is challenging in the case of
trees because it is not straightforward to identify which processes owns the
departure points.
Indeed, complications arise because not all departure points will reside in the
domain owned by the current process. Moreover, due to the irregular shapes of
the partitions, one cannot even ensure they are entirely owned by neighboring
processes.
At best we can only expect that their locations are bounded by a halo of width
$w \le \text{CFL} \: \Delta x_{\min}$ around the local partition, where
$x_{\min}$ is the size of the smallest cell in the forest.
Naturally, if one enforces $\text{CFL} \le 1$, one can ensure that the halo is
bounded by the ghost layer, which significantly simplifies the communication
problem.
This assumption, however, defeats the purpose of using a semi-Lagrangian
approach, whose purpose is to enable large CFL values.

One remedy to this problem, proposed in \cite{Thomas;Cote:95:Massively-parallel-s} for uniform grids, is to increase the size of ghost layer to $\lceil \text{CFL} \rceil$. For large values of the $\text{CFL}$ number, however, this approach can substantially increase the communication volume. Moreover, this simple approach does not work in the process of generating $G^{n+1}$ due to repartitioning. Indeed, $G^{n+1}$ is built iteratively and load balancing is enforced by repartitioning at each sub-iteration. Therefore, after one such sub-iteration, the backtracked points can end up outside of the initial ghost layer. An alternative approach would be to handle local and remote interpolations separately. Our remote interpolation algorithm is composed of three separate phases. In the first phase, which we call buffering, every process searches for all departure points inside the local trees. If the point is owned by a local cell, it is added to a local buffer, otherwise we find the process which owns the point and add the point to a separate buffer belonging to the found rank. Note that searching the point in the local tree is performed recursively using the hierarchal reconstruction (cf.\ algorithm \ref{alg:reconstruction}). Moreover, the owner's rank is found by computing the Z-index of the point and then using a binary search on the Z-curve. This is already implemented in \texttt{p4est} and explained in details in section 2.5 of \cite{Burstedde;Wilcox;Ghattas:11:p4est:-Scalable-Algo}.

@@

	\If {$\mathit{it} \not= \mathit{buff[mpirank]}.\texttt{end()}$}

@@

\begin{figure}[htbp]
\begin{center}
\begin{tabular}{ccccc}
\includegraphics[width=.28\textwidth]{figures/enright_blue_0.png}
&
\hspace{.1cm}
&
\includegraphics[width=.28\textwidth]{figures/enright_blue_1.png}
&
\hspace{.1cm}
&
\includegraphics[width=.28\textwidth]{figures/enright_blue_2.png}
\end{tabular}
\caption{Visualization of the deformation undergone by the sphere for the Enright's test at time 0, $t_{1/2}$ and $t_f$ and colored with the processes ranks. The forest's finest resolution is equivalent to a uniform grid with $2048^3$ cells and contains between 17 (at initial and final times) and 63 (at $t_{1/2}$) million nodes. These results were obtained on the Comet supercomputer with 960 processes (40 compute nodes).} \label{fig:accuracy}
\end{center}
\end{figure}

>>>>

In \texttt{p4est} the adaptive grid is represented as a non-overlapping collection of trees that are rooted in individual cells of a common coarse grid (cf.\ Figure \ref{fig:p4est_zcurve}).
This common coarse grid, which we will refer to as the ``macromesh'', can in
general be an unstructured quadrilateral mesh, in two spatial dimensions, or
hexahedral mesh, in three spatial dimensions.
(In \cite{StewartEdwards04} such a coarsest possible grid has been introduced
as the ``genesis'' mesh.)
%
In this article, however, it is sufficient to limit discussions to simple
uniform Cartesian macromeshes.
Moreover, it is implicitly assumed that the macromesh is small enough that it
can be entirely replicated on all processes.
For instance, in many of the applications that we are considering in this paper the macromesh is simply a single cubical cell.
\texttt{p4est} allows for arbitrary refinement and coarsening criteria through defining callback functions.
In this article the refinement criteria is chosen based on the distance of individual cells to the interface.
Specifically, a cell $C$ is marked for refinement if
\be
\min_{v \in V(C)} |\phi (v)| \le \frac{L D}{2},
\label{eq:refine}
\ee
where $V(C)$ denotes the set of all vertices of cell $C$, $L$ denotes the
Lipschitz constant of the level-set function, and $D$ denotes the diagonal size
of cell $C$.
Conversely, an existing cell is marked for coarsening if
\be
\min_{v \in V(C)} |\phi (v)| > L D.
\label{eq:coarsen}
\ee
We refer to section 3.2 of
\cite{Burstedde;Wilcox;Ghattas:11:p4est:-Scalable-Algo} for details on the
parallel refinement and coarsening algorithms implemented in \texttt{p4est}.

Once the grid is adapted to the interface, it must be partitioned to ensure
load balancing across processes.
This is achieved by constructing a Z-curve that traverses the leaves of all
trees in order of the tree index (cf.\ Figure \ref{fig:p4est_zcurve}).
A Z-curve is a Space Filling Curve (SFC) with the important property that cells
with close Z-indices are also geometrically close (on average) in the physical
domain.
This is beneficial since it leads to both a reduction in MPI communications and
improvements of the cache performance of several algorithms such as
interpolation and finite difference calculations.
For more details on parallel partitioning in \texttt{p4est} one may refer to
section 3.3 of \cite{Burstedde;Wilcox;Ghattas:11:p4est:-Scalable-Algo}.
\begin{figure}[hbtp]
\begin{center}
\includegraphics[width = \columnwidth]{figures/p4est_zcurve.pdf}
\caption{Left: a ``forest'' made up of two trees $T_0$ and $T_1$. Parallel
partitioning is achieved by first constructing a Z-curve starting at cell ``S''
and ending at cell ``E''.  Next, the one dimensional curve is split up among
processes either uniformly or by assigning different weights to cells.  Here
processes are represented via different colors.  Note how using the Z-curve
naturally leads to clustering of most cells in each domain.  Right: schematic of
a tree data structure representing this forest and its partitioning.%
% (The \pforest code actually supports both 2D and 3D meshes whose trees are
% rotated against each other.)
}
\label{fig:p4est_zcurve}
\end{center}
\end{figure}
Aside from grid manipulation and partitioning, we use two additional features
of \texttt{p4est}, namely the generation of ghost layer cells and the creation
of a globally unique node indexing.
These algorithms are detailed in sections 3.5 and 3.6 of
\cite{Burstedde;Wilcox;Ghattas:11:p4est:-Scalable-Algo}.
We have specifically extended the latter algorithm such that it can be applied
to a non-graded refinement pattern.
This is important because we can entirely skip the 2:1 balance function, which
was shown to be one of the more time consuming parts of grid adaptation
in \texttt{p4est}
\cite{Burstedde;Wilcox;Ghattas:11:p4est:-Scalable-Algo}.
% \cite{IsaacBursteddeGhattas12}.
% (it has since been optimized \cite{IsaacBursteddeGhattas12}).

Finally, in \texttt{p4est} trees are linearized, i.e.\ only the leaves are explicitly stored. However, explicit knowledge of the hierarchal structure of the tree is greatly beneficial in several algorithms, e.g.\ in search operations needed for the interpolation algorithm. Thus, we introduce a simple reconstruction algorithm that recreates a local representation of the entire ``forest'' that is only adapted to local cells and, potentially, the ghost layer. This approach is similar to the ideas introduced in \cite{Bangerth;Burstedde;Heister;etal:11:Algorithms-and-data-} and our tests show that in a typical application they amount to less that $1\%$ of the entire runtime. Algorithm \ref{alg:reconstruction} illustrates how this reconstruction is performed. Given a forest and a layer of ghost cells from \texttt{p4est}, the algorithm generates a local representation of the forest by recursively refining from the root until reaching the same level and location of all leaves in the local forest and the ghost layer. Note that algorithm \ref{alg:reconstruction} does not involve any communication and is load balanced provided that the initial forest is balanced. Figure \ref{fig:reconstruction} illustrates an example where Algorithm \ref{alg:reconstruction} is applied. Note how each process has independently generated a local representation of the forest that is refined to match the same leaves as in the global forest and ghost layer.
\begin{figure}[htbp]
\begin{center}
\includegraphics[width = \textwidth]{figures/reconstruct.pdf}
\end{center}
\caption{Left: a forest refined close to an interface and partitioned among four processes, as indicated by colors. Right: each process independently recreates a local forest that is refined to match the local grid and is as coarse as possible elsewhere. Note that empty cells are fictitious, i.e.\ they are only required to generate the hierarchal structure and are not matched by any corresponding cell in the global forest.}
\label{fig:reconstruction}
\end{figure}
\begin{algorithm}[htbp]
\caption{$H \gets \texttt{Reconstruct (}G\texttt{)}$: 
Construction of the local tree hierarchy, $H$, from the parallel grid, $G$, supplied by \texttt{p4est}. The algorithm starts at the coarsest level, i.e. the macromesh, and recursively splits the each cell to match the finest local and ghost cells generated by \texttt{p4est}.}
\begin{algorithmic}[1]
\State $H \gets G.\texttt{macromesh()}$ \Comment{start with the coarsest  grid}
\For {$\mathit{tr} \ein G.\texttt{local\_trees()}$} \Comment{update the hierarchy to match local cells}
	\For {$c \ein \mathit{tr}.\texttt{local\_cells()}$}
		\State $H.\texttt{update\_tree(} \mathit{tr}, c \texttt{)}$
	\EndFor
\EndFor

@@

\subsection{Interpolation and semi-Lagrangian methods} \label{sec:interpolation_semilagrangian}
As indicated earlier, we use the semi-Lagrangian method to solve equation \eqref{eq:ls} when the velocity field is externally generated, i.e.\ when it does not depend explicitly on the level-set function itself. Let us rewrite equation \eqref{eq:ls} along the characteristic curve $\underline{\mathbf{X}}(t)$ as:
\be
\left\{
\begin{array}{rcl}
\dfrac{\text{d} \underline{\mathbf{X}}}{\text{d} t} &=& \underline{\mbf{u}}, \\ [3ex]
\dfrac{\text{d} \phi(\underline{\mathbf{X}}(t), t)}{\text{d} t} &=& 0.
\end{array}
\right.
\label{eq:sl}
\ee
The semi-Lagrangian method integrates equations \eqref{eq:sl} backward in time,
i.e.\ starting from the grid $G^{n+1}$ (computed iteratively as explained
later on), we simply write $\phi^{n+1}(\underline{\mathbf{X}}^{n+1}) =
\phi(\underline{\mathbf{X}}(t^{n+1}), t^{n+1}) =
\phi(\underline{\mathbf{X}}(t^n), t^n) = \phi^n(\underline{\mathbf{X}}_d)$.
Here, the characteristic curves are chosen such that $\underline{\mathbf{X}}(t^{n+1})$ are the coordinates of grid points of $G^{n+1}$, and $\underline{\mathbf{X}}_d$ are the departure points, which are computed using the second-order midpoint method \cite{Min;Gibou:07:A-second-order-accur}:
\bea
\underline{\mathbf{X}}^\star &=& \underline{\mathbf{X}}^{n+1} - \frac{\Delta t}{2} \underline{\mbf{u}}^{n}(\underline{\mathbf{X}}^n),	   \label{eq:xstar}      \\
\underline{\mathbf{X}}_d     &=& \underline{\mathbf{X}}^{n+1} - \Delta t \underline{\mbf{u}}^{n+\frac{1}{2}}(\underline{\mathbf{X}}^\star), \label{eq:xdeparture}
\eea
where $\underline{\mbf{u}}^{n+\frac{1}{2}}$ is obtained via extrapolation from previous times, i.e.:
\be
\underline{\mbf{u}}^{n+\frac{1}{2}} = \frac{3}{2} \underline{\mbf{u}}^n - \frac{1}{2}\underline{\mbf{u}}^{n-1}. \label{eq:vn_p_half}
\ee
Note that all values at the intermediate point, $\underline{\mathbf{X}}^\star$,
and departure point, $\underline{\mathbf{X}}_d$, must be calculated via
interpolation from the previous grids $G^{n}$ and $G^{n-1}$.
Here, we use the stabilized second-order interpolation for
$\phi(\underline{\mathbf{X}}_d)$ and the multi-linear interpolation for
$\underline{\mbf{u}}^{n+\frac{1}{2}}(\underline{\mathbf{X}}^\star)$
\cite{Min;Gibou:07:A-second-order-accur}.
Although parallelization of the interpolation process on a shared-memory
machine is trivial, the same cannot be said for distributed-memory machines.
In fact, the parallel interpolation given in Algorithm \ref{alg:interpolation}
is probably the most important contribution of this article since this
procedure, which is trivial on uniform grids, is challenging in the case of
trees because it is not straightforward to identify which processes owns the
departure points.
Indeed, complications arise because not all departure points will reside in the
domain owned by the current process. Moreover, due to the irregular shapes of
the partitions, one cannot even ensure they are entirely owned by neighboring
processes.
At best we can only expect that their locations are bounded by a halo of width
$w \le \text{CFL} \: \Delta x_{\min}$ around the local partition, where
$x_{\min}$ is the size of the smallest cell in the forest.
Naturally, if one enforces $\text{CFL} \le 1$, one can ensure that the halo is
bounded by the ghost layer, which significantly simplifies the communication
problem.
This assumption, however, defeats the purpose of using a semi-Lagrangian
approach, whose purpose is to enable large CFL values.

One remedy to this problem, proposed in \cite{Thomas;Cote:95:Massively-parallel-s} for uniform grids, is to increase the size of ghost layer to $\lceil \text{CFL} \rceil$. For large values of the $\text{CFL}$ number, however, this approach can substantially increase the communication volume. Moreover, this simple approach does not work in the process of generating $G^{n+1}$ due to repartitioning. Indeed, $G^{n+1}$ is built iteratively and load balancing is enforced by repartitioning at each sub-iteration. Therefore, after one such sub-iteration, the backtracked points can end up outside of the initial ghost layer. An alternative approach would be to handle local and remote interpolations separately. Our remote interpolation algorithm is composed of three separate phases. In the first phase, which we call buffering, every process searches for all departure points inside the local trees. If the point is owned by a local cell, it is added to a local buffer, otherwise we find the process which owns the point and add the point to a separate buffer belonging to the found rank. Note that searching the point in the local tree is performed recursively using the hierarchal reconstruction (cf.\ Algorithm~\ref{alg:reconstruction}). Moreover, the owner's rank is found by computing the Z-index of the point and then using a binary search on the Z-curve. This is already implemented in \texttt{p4est} and explained in details in section 2.5 of \cite{Burstedde;Wilcox;Ghattas:11:p4est:-Scalable-Algo}.

@@

	\If {$\mathit{it} \not= \mathit{local\_buff[mpirank]}.\texttt{end()}$}

@@

\begin{figure}[htbp]
\begin{center}
\begin{tabular}{ccccc}
\includegraphics[width=.28\textwidth]{figures/enright_blue_0.png}
&
\hspace{.1cm}
&
\includegraphics[width=.28\textwidth]{figures/enright_blue_1.png}
&
\hspace{.1cm}
&
\includegraphics[width=.28\textwidth]{figures/enright_blue_2.png}
\end{tabular}
\caption{Visualization of the deformation undergone by the sphere for the Enright's test at time 0, $t_{1/2}$ and $t_f$. The forest's finest resolution is equivalent to a uniform grid with $2048^3$ cells and contains between 17 (at initial and final times) and 63 (at $t_{1/2}$) million nodes. These results were obtained on the Comet supercomputer with 960 processes (40 compute nodes).} \label{fig:accuracy}
\end{center}
\end{figure}
